---
title: "Introduction to R"
output:
  html_document:
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, fig.width = 5, fig.height = 5, fig.align = 'center', eval = T, collapse = T)
```

# The R language

R is an interpreted language, which means that most of its implementations execute instructions directly and freely, without previously compiling a program into machine-language instructions. Thus, users can typically access R through a command-line interpreter, such as conda's command prompt (in Windows) or the terminal (in GNU/Linux). For instance, we can type `2+2`, press <kbd>Enter</kbd> (in a terminal) or <kbd>Ctrl</kbd>+<kbd>Enter</kbd> (in RStudio), the computer replies 4.

```{r, eval=TRUE}
2 + 2
```

# Objects

In R, *objects* provide a means of accessing data stored in memory. Objects are referred to through *symbols* (variable names) or *variables*. 

## Vectors

Vectors are the most basic objects in R. They can be thought of as contiguous cells containing data. Single numbers, such as `3.14`, and strings, such as `"pi"`, are still vectors (of length 1). Vectors of length 0 are also possible, and often useful.

R has six basic *atomic* vector types, where atomic means all their elements are of the same type. Those basic types are the following.

### Logical

Takes two possible values: `TRUE` (`T`) or `FALSE` (`F`).
```{r, eval=T}
var_logical = TRUE
typeof(var_logical)

vector_logical = c(TRUE, FALSE, TRUE)
typeof(vector_logical)

vector_logical[1]

is.logical(vector_logical)

vector_logical[4]
is.logical(vector_logical[4])

```
`NA`(not available) is a special type of data (by default logical).

### Integer

```{r, eval=T}
var_int = 2L
typeof(var_int)

is.integer(var_int)

length(var_int)

var_int
var_int[1]

vector_int = c(var_int, 3L, 8L, 13:18)
typeof(vector_int)

vector_int

length(vector_int)

vector_int[10]

typeof(vector_int[10])
```

### Double (numeric)

```{r, eval=T}
var_double = 3.14
typeof(var_double)
is.double(var_double)
is.numeric(var_double)

var_inf = var_double/0; var_inf
var_nan = 0/0; var_nan
typeof(c(var_inf, var_nan))
```
`Inf` and `NaN` are special types of double.

### Complex

```{r, eval=T}
a = 5.2
b = 1.7
c = complex(real = a, imaginary = b); c
typeof(c)
Re(c)
Im(c)
Mod(c)
Arg(c)
Conj(c)
sqrt(c)
```
### Character

```{r, eval=T}
var_chr1 = "Hola"
var_chr2 = "Mundo"

typeof(var_chr1)

c(var_chr1, var_chr2)

paste(var_chr1, var_chr2)
```

### Raw 
```{r, eval=T}
var_raw = as.raw(40); var_raw
```
Notice the hexadecimal output.
```{r, eval=T}
vect_raw = charToRaw("Hola Mundo"); vect_raw

rawToChar(vect_raw[1:4])

typeof(vect_raw[1:4])
```

## Lists

Lists are another kind of data storage. Lists have elements, but unlike vectors, those elements do not have to be of the same type. Here are some examples.

```{r, eval=T}
my.contact = list(
  name = "Juan C",
  gitpage = "http://github.com/jlaria",
  age = 26,
  tags = c("#data-science", "#r", "#python")
)
my.contact
```
We can access specific element of the list with `$`
```{r, eval=T}
my.contact$tags
typeof(my.contact$tags)
```

## Factors

Factors are special compound objects, used to describe cathegorical data.

```{r, eval=TRUE}
risk = factor(c("hi", "hi", "lo", "med", "lo", "hi")); risk

typeof(risk)

levels(risk)

risk = factor(c("hi", "hi", "lo", "med", "lo", "hi"),
              levels=c("lo", "med", "hi"),
              ordered = T); risk
```

## Matrices

A matrix is a collection of similar object arranged in rows and columns. Matrices are constructed column-wise so entries start in the upper left corner and run down the columns.

```{r}
X = matrix(1:8, nrow = 2, ncol = 4); X

dim(X)

t(X)

norm(X, "F")

rownames(X) = c("obs 1", "obs 2")
colnames(X) = paste("var",1:4)
X
```

Matrices can also be created by column-binding or row-binding.

```{r}
x = 1:4
y = 5:8

cbind(x,y)

rbind(x,y)
```

Multi-dimensional arrays are also posible.

```{r}
A = array(1:24, dim = c(2,4,3)); A
```


# Indexing

R contains several constructs which allow access to individual elements or subsets through indexing operations.

There are three basic indexing operators: `[]`, `[[]]` and `$`. 


```{r, eval=T}
vector_int

vector_int[9]

vector_int[c(2,9)]

X[2,1]

X["obs 2", "var 1"]

A[1,2:3,1:2]
```
Components of vectors can also be named.
```{r, eval=T, collapse=T}
namedvector_int = c(1, 1, 2, 3, 5)
names(namedvector_int) = c("day 1", "day 2", "day 3", "day 4", "day 5")

namedvector_int["day 1"]

namedvector_int[c("day 5", "day 2")]
```

The operator `[[]]` is useful with list vectors. It can be used on atomics, tho.

```{r, eval=T}
my.contact[4] # Returns a list with one element: a character vector

typeof(my.contact[4])

my.contact[[1]] # Returns the character vector 

typeof(my.contact[[1]])
```

The form `$` applies only to lists objects.

```{r, eval=T}
my.contact$name

my.contact$"age"
```

It is also possible to index vectors with `[]` using more complex constructions.

```{r}
a = c(1, 3, 4, -1, 2)
b = c(1, 2, 4, 8, 16)

a[b]

b[-4]

a[c(T,F,T)]

b[risk]

b[]
```

# Operators

R allows the use of arithmetic expressions using operators, which can be grouped using parentheses, mixed with function calls, and assigned to variables in a straightforward manner.

```{r}
x = 2; a = 1
y = 2 * (a + log(x)); y
```

Common vector/matrix opperators are listed here.

  `-`	Minus, can be unary or binary
```{r}
x = c(1,-1, 3)
y = c(2, 4, 6)

x-y
```
  
  
  `+`	Plus, can be unary or binary
```{r}
x+y
```
  
  `!`	Unary not
```{r}
b = c(T, F, T)

!b
```
  
  `:`	Sequence, binary (in model formulae: interaction)
```{r}
z = 3:8; z
```
  
  `*`	Multiplication, binary
```{r}
x*y

2*z

x*z
```

  `/`	Division, binary

```{r}
x/y
```


  `^`	Exponentiation, binary
```{r}
x^y
```

  `%%`	Modulus, binary
```{r}
7%%5
```

  `%/%`	Integer divide, binary
```{r}
7%/%5
```

  `%*%`	Matrix product, binary
```{r}
x %*% y
```

  `%o%`	Outer product, binary
```{r}
M = x %o% y; M

```

  `%x%`	Kronecker product, binary
```{r}
x %x% M
```

  `%in%`	Matching operator, binary (in model formulae: nesting)
  
```{r}
x %in% z
```

  `<`	Less than, binary

```{r}
x < y
```


  `>`	Greater than, binary

```{r}
y > x
```

  `==`	Equal to, binary

```{r}
x == y
```

  `>=`	Greater than or equal to, binary

  `<=`	Less than or equal to, binary

  `&`	And, binary, vectorized
  
```{r}
c(T, F, T) & c(F, F, T)
```
  

  `&&`	And, binary, not vectorized
```{r}
c(T, F, T) && c(F, F, T)
```

  `|`	Or, binary, vectorized
```{r}
c(T, F, T) | c(F, F, T)
```

  `||`	Or, binary, not vectorized
  
```{r}
c(T, F, T) || c(F, F, T)
```

  `<-`	Left assignment, binary
```{r}
a <- 1; a
```

  `->`	Right assignment, binary
```{r}
1 -> a; a


```

# Control structures

Computation in R consists of sequentially evaluating **statements**. Statements, such as `a = 1` or `a^2`, can be separated by either `;` or a new line. They can also be grouped together using braces `{` and `}`, forming **blocks**.

## if

The `if/else` statement conditionally evaluates two statements. There is a `condition` which is evaluated and if the value is `TRUE` then the first statement is evaluated; otherwise the second statement will be evaluated.

```{r, eval=F}
if(condition){
  statement 1
}else{
  statement 2
}
```

## Loops

R has three statements that provide explicit looping. They are `for`, `while` and `repeat`. The two built-in constructs, `next` and `break`, provide additional control over the evaluation. R provides other functions for implicit looping such as `tapply`, `apply`, and `lapply`. In addition many operations, especially arithmetic ones, are vectorized so you may not need to use a loop. 

If there is some other way, avoid using loops in R, the former is probably faster.

### for

The syntax of the `for` loop is
```{r, eval=F}
for(object in vector){
  statement
}
```

Notice that if `statement` is only one line, braces `{}` are not required. 

### while

The syntax of the `while` loop is
```{r, eval=F}
while(condition){
  statement
}
```

### repeat 

The syntax of the `repeat` loop is 

```{r, eval=F}
repeat{
  statement
}
```

The `repeat` statement causes repeated evaluation of the body until a `break` is specifically requested. This means that you need to be careful when using `repeat` because of the danger of an infinite loop.

## Implicit loops

The `apply` family of fuctions avoids the explicit use of loop constructs. They manipulate slices of data from matrices, arrays, lists and dataframes in a repetitive way. Here are some important members of this family.

### apply

```{r}
X = matrix(rnorm(4*6), nrow = 4, ncol = 6); X

# We want to compute the mean of each column

col.means = apply(X, 2, mean); col.means

# col.means is a vector and it is equivalent to
col.means_for = rep(0, 6) # declare it
for(j in 1:6){
  col.means_for[j] = mean(X[,j])
}
col.means_for
```

# Data manipulation

## Data frames

Data frames in R can be seen as lists of vectors, all with the same length. They are by far the most useful storage structure for data analysis in R. 
Data frames group vectors together into a 2D tables, where each vector is a column of the table. As happpens with lists in R, each column of a data frame can contain a different type of data, but since columns are vectors, every cell within a column must be of the same type.

There are many ways to create data frames in R, but we rarely have to type the data in R.

```{r}
df = data.frame(name = c("Juan", "Pedro", "María"),
                age = c(26, 33, 22),
                phone = c("666-849334", "666-837489", "666-029340") )
df
```

We can create a data frame from a previously defined 2D data matrix.

```{r}
X = matrix(1:8, nrow = 2, ncol = 4)

df = data.frame(X)
df
```

We can also assign names to columns and rows

```{r}
colnames(df) = paste0("Col", 1:4)
rownames(df) = paste0("Obs", 1:2)

df
```

We can delete and edit columns, rows, and cells.

```{r}
df = data.frame(matrix(rnorm(40), nrow=5, ncol=8))

df$X1 = NULL # delete column 1
df

df[,c(3,6)] = NULL # delete columns 3 and 6
df

df = df[-1, ] # remove row 1
df

df = df[,-c(4:6)] # remove columns 4,5,6
df

df[2,] = -10 # edit row 2
df

df$X5 = 5 # edit column X5
df

```

Data frames with the same colums can be row-concatenated with `rbind`.

```{r}
df1 = data.frame(name = c("Pedro", "María"),
                age = c(33, 22),
                phone = c("666-837489", "666-029340") )
df2 = data.frame(name = c("Juan"),
                age = c(26),
                phone = c("666-849334") )

rbind(df1, df2)
```

Data frames with the same rows can be column-concatenated with `cbind`.
```{r}
df1 = data.frame(name = c("Juan", "Pedro", "María"),
                age = c(26, 33, 22))
df2 = data.frame( phone = c("666-849334", "666-837489", "666-029340") )

cbind(df1, df2)
```

## Data input

### The working directory

Every R session has a default location on the operating system’s file structure called the working directory.

The fuction `getwd()` tells what the current working directory is.

```{r}
getwd()
```

To change the working directory, use the `setwd` function. Be sure to enter the working directory as a character string (enclose it in quotes).

```{r, eval=F}
setwd("~/juank/Downloads")
```

### Reading data

The `read.table` function reads a file in table format and automatically creates a data frame from it, with cases corresponding to rows and variables to columns in the file.

```{r}
df = read.table("data/protein.csv",
                header = T,
                sep = ",")
head(df)
```

The root path to read and write data is the current working directory, however we can specify the whole path to the data. Some reading functions allow to read from urls, not only local paths.

When reading files from the internet, it is often convenient to use the `nrows` parameter the first time, with a reasonable value, to check if we are getting the format right.

```{r}
df = read.table("http://hci.stanford.edu/courses/cs448b/data/ipasn/cs448b_ipasn.csv", nrows = 10)
df
```

For instance, in this case, the data is a standard comma separated `.csv` file. We can set the parameter `sep=,` or just use the function `read.csv`.

```{r}
df = read.csv("http://hci.stanford.edu/courses/cs448b/data/ipasn/cs448b_ipasn.csv", nrows = 10)
df
```

If you are working in a desktop computer and your file is in a local directory, you can use the function `file.choose` combined with `read.table` or `read.csv`.

```{r, eval=F}
df = read.csv(file.choose())
```

### Writing data

To export data from R, we can use the `write.table` and `write.csv` functions. Notice that R will export the data to the current directory if we do not specify a root path.

# Functions

## Scope

Almost every programming language has a set of scoping rules, allowing the same name to be used for different objects. This allows, e.g., a local variable in a function to have the same name as a global object. 

The **global environment** is the root of the user workspace. An assignment operation from the command line will cause the relevant object to belong to the global environment. 

If we start up R and make an assignment, we are adding an entry in the global environment.

```{r}
a = 1234

ls()  # to list all the object in the (global) environment
```

Here we can see all the variables that we have defined earlier in this document, including `a`.
We can think of environments as boxes that store objects, including other boxes.

We can define a new environment with the function `new.env`, and create variables inside that environment with `assign`.

```{r}
my.env = new.env()
assign("a", 123, envir = my.env)
ls(my.env)
my.env$a
```

The `$` form allows to create, read, and modify an object inside an environment.

```{r}
my.env$a = my.env$a*2
my.env$a
```

Objects in R can be deleted with the function `rm`. 

```{r}
rm(my.env) # deletes the environment my.env (including all the objects inside that enviroment)
```

We can not delete the global environment, but we can delete all the objects defined inside.

```{r}
rm(list = ls()) # deletes everything in the global environment

ls() # empty
```

Functions have their own environments. 

## Function syntax

The syntax for writing a function is 

```{r, eval=F}
f = function(args){
  body
}
```

The first component of the function declaration is the keyword `function` which indicates to R that you want to create a function.

An argument list is a comma separated list of formal arguments. A formal argument can be a symbol, a statement of the form `symbol = expression`, or the special formal argument `...`

Generally functions are assigned to symbols but they don’t need to be. The value returned by the call to function is a function. If this is not given a name it is referred to as an `anonymous` function. Anonymous functions are most frequently used as arguments to other functions such as the `apply` family. 

The return value of the function can be specified with `return`.

```{r}
add = function(x, y){
  return(x+y)
}
add(2,2)
```

If the `return` value is not specified, by default the value of the last evaluation will be returned.

```{r}
add = function(x, y){
  x+y
}
add(2,2)
```

For example, we can create a function that writes the date followed by the value of some variable.
```{r}
wlog = function(x){
  write(paste0(date(), "\t", x), "")
}

x = 3.14
wlog("Hola Mundo")
wlog(x)

```

Furthermore, we can specify default values.

```{r}
wlog = function(x, y=5){
  write(paste0(date(), "\t", x, "\t", y), "")
}

wlog(1.2)
wlog(1.2, 2.4)
```

We can specify with the default value `NULL` that a particular argument is optional.

```{r}
wlog = function(x, y=NULL){
  write(paste0(date(), "\t", x, "\t", y), "")
}

wlog(1.2)
wlog(1.2, 2.4)
```
The argument `...` can be passed to function calls inside our function, meaning an arbitrary number of arguments.

```{r}
wlog = function(..., printdate=T){
  write(paste0( 
      {if(printdate)date() else NULL},
      ...), "")
}
wlog("\tHola mundo")
wlog(printdate = F, "Hola Mundo!")
wlog("\t x = ",2, "\t y = ", 5, "\t x + y = ", 2+5, "\t FIN.")

```


